Constructing Quotient Inductive-Inductive Types

### 2018-10-10 paper.tex
fixes suggested by first reviewer

In the rest of the introduction, before giving an overview of the
paper, we illustrate our method for deriving the above notions by
three examples. We start with a closed type which has a recursive
constructor, the natural numbers. Then we move on to a parameterised
type with an equality constructor: the integers. Finally, we sketch
how our method works for the above $\Con$-$\Ty$ example. % A more formal
% discussion of the metatheory and signatures will follow in Section
% \ref{sec:metatheory} and \ref{sec:signatures}.

%% Our notationn
%% for the metatheory is standard, but we summarise it in Section
%% \ref{sec:metatheory}.

@@

Given a displayed algebra $(N^D,z^D,s^D)$ over the initial algebra $\con_\Delta$,
we construct a section which we call the eliminator. It has type
$\Delta^\S\,\con_\Delta\,(N^D,z^D,s^D)$ --- the two equations in the
definition of sections correspond to the $\beta$-rules.
\begin{alignat*}{5}
  & \elim_\Delta\,(N^D,z^D,s^D) && \equiv \big( && \lambda t.\tr_{N^D}\,({t^\C\,\id}^{-1})\,(t^\D\,(N^D,z^D,s^D)),\, \refl_{z^D}, \\
  & && && \lambda x\,x^D. \J\,\refl_{s^D\,x\,(x^\D\,(N^D,z^D,s^D))}\,(x^\C\,\id)\big)
\end{alignat*}
We can eliminate any natural number $t$ using the logical predicate
interpretation $t^\D\,(N^D,z^D,s^D): N^D\,(t^\A\,\con_\Delta)$. The
result has to be transported along the equality $t^\C\,\id : (t =
t^\A\,\con_\Delta)$ so that we get something of type $N^D\,t$. The
operation $\blank^\C$ is a generalisation of $\con$; as we will see
later, $\con$ is defined in terms of $\blank^\C$. The usage of
$t^\C\,\id$ corresponds to the identity extension lemma
\cite{atkey}. The computation rule for zero is strict, but the case
for successor requires using $\J$ on $x^\C\,\id$.

@@

% We give an overview of works describing schemes for various classes of
% inductive types.
Internal codes for simple inductive types such as
natural numbers, lists or binary trees can be given by containers
which are decoded to W-types \cite{abbot05containers}. Morris and
Altenkirch \cite{morris09indexed} extend the notion of container to
that of indexed container which specifies indexed inductive
types. External schemes for inductive families are given in
\cite{Dybjer97inductivefamilies,paulinmohring}, for
inductive-recursive types in \cite{dybjer00ir}. Inductive-inductive
types were introduced by Forsberg together with an internal coding
scheme \cite{forsberg-phd}. A symmetric scheme for both inductive and
coinductive types is given in \cite{henning}.

Quotient types as in \cite{hofmann95extensional} are a precursor to
the current development. More recently, an increasing amount of
research is concerned with higher inductive types (HITs), motivated by
their use cases in homotopy type theory \cite{HoTTbook}. The theory of
HITs is relevant to the current work as it also has to describe
signatures with equality constructors and algebras with
equalities. \cite{niels} define an external syntactic scheme for
higher inductive types with only 0-constructors and compute the types
of elimination principles. In \cite{nielsmsc} a semantics is given for
the same class of HITs but with no recursive equality
constructors. Sojakova \cite{sojakova} defines a subset of HITs
called W-suspensions by an internal coding scheme similar to
W-types. She proves that the induction principle is equivalent to
homotopy initiality. Dybjer and Moeneclaey define a syntactic scheme
for a some HITs
and show their existence in a groupoid model
\cite{moeneclaey}. In \cite{AwodeyFS18}, impredicative encodings of a
class of higher inductive types are presented, and we think that this
approach could also be modelled within our framework once we assume an
impredicative universe.

@@

Our metatheory is Martin-L{\"o}f type theory with functional
extensionality and uniqueness of identity proofs (UIP). In this
section we describe the notation in this paper and the accompanying
Agda formalisation.

Definitional equality is denoted by $\equiv$. We have a cumulative
hierarchy of Russell-style universes $\Set_i$ where we usually omit
indices. Dependent function space is denoted $(\alpha:T)\ra T'$. $T\ra
T'$ stands for $(\alpha:T)\ra T'$ if $T'$ does not depend on
$\alpha$. We use $(\alpha:T)(\alpha': T')\ra T''$ as a shorthand for
iterated function spaces. We sometimes omit certain arguments of
functions or write them in subscript to lighten the notation.
$\Sigma$ types are denoted by $(\alpha:T)\times T'$ with
left-associative $\times$. The constructor is $(\blank,\blank)$ and
the eliminators are written $\proj_1$ and $\proj_2$. The one-element
type $\top$ has constructor $\tt$. We have the identity type $=$ with
constructor $\refl$, eliminator $\J$. The notation is
$\J_{x.z.P}\,pr\,e:P[x\mapsto\alpha',z\mapsto e]$ for
$\alpha,\alpha':T$ and $x:T, z:\alpha=x \vdash P:\Set$ and
$pr:P[x\mapsto \alpha,z\mapsto\refl]$ and $e :\alpha=\alpha'$. We
write $\tr_P\,e\,u: P\,\alpha'$ for transport of $u : P\,\alpha$ along
$e :\alpha=\alpha'$. Sometimes we omit the parameters in subscript. We
also have coercion $\coe\,e\,\alpha:T'$ whenever $e:T=T'$ and the
inverse $e^{-1}:\alpha'=\alpha$ for $e:\alpha=\alpha'$.

In this paper we use the notation of extensional type theory, that is,
after proving an an equality $p : \alpha=\alpha'$, in later proofs
depending on this equality we treat it as definitional, thus
$\alpha\equiv\alpha'$ and $p \equiv refl$. This makes the notation
much lighter, as transports disappear: $\tr\,p\,u\equiv u$ for any
proof $p$. Although we use extensional notation, this can
be translated to intensional type theory extended with functional
extensionality and UIP, as we know from
\cite{hofmann95conservativity, Oury2005,theo}. In our Agda formalisation, we use these axioms
along with rewrite rules \cite{cockx}. A
rewrite rule allows to turn a previously proven (or postulated)
equality into a definitional one. % Whenever Agda encounters a stuck
% term, it looks at the current list of postulated rewrite rules and
% checks whether the term matches the left hand side of one of them. If
% this is the case, it rewrites the term along that rule.
This provides
an alternative way to add inductive types to Agda: one can postulate
the constructors, the eliminator and the computation rules and make
the computation rules definitional by marking them as rewrite rules.

We denote the functional extensionality axiom by $\funext : (\alpha :
A)\ra f\,\alpha = g\,\alpha \ra f = g$. We also use the term $\UIP$
which has type $e = e'$ whenever $e,e':\alpha=\beta$. We write
equational reasoning proofs by writing the proofs of equalities above
the $=$ symbol, e.g. $u \overset{e}{=} u'$ when $e:\alpha = \alpha'$
and $u:P\,\alpha$, $u':P\,\alpha'$.

@@

As an example of using the susbtitution calculus, we prove the functor
laws for terms using equational reasoning.
\begin{alignat*}{15}
  & [\id] && : && t[\id] &&                      && && \hspace{6em} [\circ] && : && t[\sigma\circ\delta] \\
  & && && && && {\rhd\beta_2}^{-1}                && && && && && {\rhd\beta_2}^{-1} \\
  & && = \,\, && \pi_2\,(\id, t[\id]) &&         && && && = \,\, && \pi_2\,(\sigma\circ\delta, t[\sigma\circ\delta]) \\
  & && && && && {\idl}^{-1}                       && && && && && {\idl}^{-1} \\
  & && = \,\, && \pi_2\,(\id\circ\id, t[\id])    && && && && = \,\, && \pi_2\,(\id\circ(\sigma\circ\delta), t[\sigma\circ\delta]) \\
  & && && && && {,\circ}^{-1}                    && && && && && {,\circ}^{-1} \\
  & && = \,\, && \pi_2\,((\id, t)\circ\id) &&    && && && = \,\, && \pi_2\,((\id, t)\circ(\sigma\circ\delta)) \\
  & && && && && \idr                             && && && && && \ass^{-1} \\
  & && = \,\, && \pi_2\,(\id, t) &&              && && && = \,\, && \pi_2\,(((\id, t)\circ\sigma)\circ\delta) \\
  & && && && && {\rhd\beta_2}                    && && && && && {,\circ} \\
  & && = \,\, && t &&                            && && && = \,\, && \pi_2\,((\id\circ\sigma, t[\sigma])\circ\delta) \\
  & && && &&                                     && && && && && && {,\circ} \\
  & && && &&                                     && && && = \,\, && \pi_2\,((\id\circ\sigma)\circ\delta, t[\sigma][\delta]) \\
  & && && &&                                     && && && && && && {\rhd\beta_2} \\
  & && && &&                                     && && && = \,\, && t[\sigma][\delta]
\end{alignat*}

@@

We start by defining $\Con$ as the type of $\mathbb{N}$-algebras,
$\Ty$ as displayed $\mathbb{N}$-algebras, $\Tms$ as
$\mathbb{N}$-homomorphisms and $\Ty$ as displayed
$\mathbb{N}$-algebras, using the same definitions as in section
\ref{sec:naturals}.

@@

\begin{figure}
{\footnotesize
\[
    \begin{tikzpicture}[
node distance = 0pt,
    BC/.style = {
        decorate,
        decoration={calligraphic brace, amplitude=1.2mm,
        pre =moveto, pre  length=0.75pt,
        post=moveto, post length=0.75pt,
        raise=1mm,
        #1},% for mirroring of brace
        very thick,
        pen colour={black},
                  },
%  BC/.default = mirror,
    LN/.style = {inner xsep=4pt, outer sep=0pt},
                        ]
\matrix (m) [matrix of nodes, inner sep=0pt,
             nodes={text depth=0.8ex, text height=1em, %minimum width=5ex,
                    inner ysep=1pt, inner xsep=4pt, outer sep=0pt, anchor=west},
             nodes in empty cells,
             column sep=-\pgflinewidth,
             row sep= -\pgflinewidth,
             ]
{
             & $\blank^\A$   & $\blank^\D$    & $\blank^\M$  & $\blank^\D$  & $\cdot_{\blank}$ & $\rhd_{\blank}$ & ... & $\K_{\blank}$ & ${\K[]}_{\blank}$ & ... & $\Eq_{\blank}$  & ... & \hspace{2em} & $\blank^\C$ & $\blank^\R$ & $\blank^\E$ &  \\
    $\Con$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Ty$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Tms$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Tm$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\cdot$  &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\rhd$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\U$     &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\El$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Pi$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Id$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Pim$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
};
\draw           (m-1-1.south west) -- (m-1-13.south east);
\draw           (m-1-15.south west) -- (m-1-17.south east);

>>>>

In the rest of the introduction, before giving an overview of the
paper, we illustrate our method for deriving the above notions by
three examples. We start with a closed type which has a recursive
constructor, the natural numbers. Then we move on to a parameterised
type with an equality constructor: the integers. Finally, we sketch
how our method works for the above $\Con$-$\Ty$ example. Our notation
below is standard, but we summarise it in Section
\ref{sec:metatheory}.

@@

Given a displayed algebra $(N^D,z^D,s^D)$ over the initial algebra $\con_\Delta$,
we construct a section which we call the eliminator. It has type
$\Delta^\S\,\con_\Delta\,(N^D,z^D,s^D)$ --- the two equations in the
definition of sections correspond to the $\beta$-rules.
\begin{alignat*}{5}
  & \elim_\Delta\,(N^D,z^D,s^D) && \equiv \big( && \lambda t.\tr_{N^D}\,({t^\C\,\id}^{-1})\,(t^\D\,(N^D,z^D,s^D)),\, \refl_{z^D}, \\
  & && && \lambda x\,x^D. \J\,\refl_{s^D\,x\,(x^\D\,(N^D,z^D,s^D))}\,(x^\C\,\id)\big)
\end{alignat*}
We can eliminate any natural number $t$ using the logical predicate
interpretation $t^\D\,(N^D,z^D,s^D): N^D\,(t^\A\,\con_\Delta)$. The
result has to be transported along the equality $t^\C\,\id : (t =
t^\A\,\con_\Delta)$ so that we get something of type $N^D\,t$. The
operation $\blank^\C$ is a generalisation of $\con$; as we will see
later, $\con$ is defined in terms of $\blank^\C$. The usage of
$t^\C\,\id$ corresponds to the identity extension lemma
\cite{atkey}. The computation rule for zero is definitional (can be
proved by $\refl$), but the case for successor requires using $\J$ on
$x^\C\,\id$.

@@

% We give an overview of works describing schemes for various classes of
% inductive types.
Internal codes for simple inductive types such as
natural numbers, lists or binary trees can be given by containers
which are decoded to W-types \cite{abbot05containers}. Morris and
Altenkirch \cite{morris09indexed} extend the notion of container to
that of indexed container which specifies indexed inductive
types. External schemes for inductive families are given in
\cite{Dybjer97inductivefamilies,paulinmohring}, for
inductive-recursive types in \cite{dybjer00ir}. Inductive-inductive
types were introduced by Nordvall Forsberg together with an internal coding
scheme \cite{forsberg-phd}. A symmetric scheme for both inductive and
coinductive types is given in \cite{henning}.

Quotient types as in \cite{hofmann95extensional} are a precursor to
the current development. More recently, an increasing amount of
research is concerned with higher inductive types (HITs), motivated by
their use cases in homotopy type theory \cite{HoTTbook}. The theory of
HITs is relevant to the current work as it also has to describe
signatures with equality constructors and algebras with
equalities. \cite{niels} define an external syntactic scheme for
higher inductive types with only 0-constructors and compute the types
of elimination principles. In \cite{nielsmsc} a semantics is given for
the same class of HITs but with no recursive equality
constructors. Sojakova \cite{sojakova} defines a subset of HITs
called W-suspensions by an internal coding scheme similar to
W-types. She proves that the induction principle is equivalent to
homotopy initiality. Dybjer and Moeneclaey define a syntactic scheme
for some HITs
and show their existence in a groupoid model
\cite{moeneclaey}. In \cite{AwodeyFS18}, impredicative encodings of a
class of higher inductive types are presented, and we think that this
approach could also be modelled within our framework once we assume an
impredicative universe.

@@

Our metatheory is Martin-L{\"o}f type theory with functional
extensionality and uniqueness of identity proofs (UIP). In this
section we describe the notation used in this paper and the accompanying
Agda formalisation.

Definitional equality is denoted by $\equiv$. We have a cumulative
hierarchy of Russell-style universes $\Set_i$ where we usually omit
indices (we don't assume any impredicativity). Dependent function
space is denoted $(\alpha:T)\ra T'$. $T\ra T'$ stands for
$(\alpha:T)\ra T'$ if $T'$ does not depend on $\alpha$. We use
$(\alpha:T)(\alpha': T')\ra T''$ as a shorthand for iterated function
spaces. We sometimes omit certain arguments of functions or write them
in subscript to lighten the notation.  $\Sigma$ types are denoted by
$(\alpha:T)\times T'$ with left-associative $\times$. The constructor
is $(\blank,\blank)$ and the eliminators are written $\proj_1$ and
$\proj_2$. The one-element type $\top$ has constructor $\tt$. We have
the identity type $=$ with constructor $\refl$, eliminator $\J$. The
notation is $\J_{x.z.P}\,pr\,e:P[x\mapsto\alpha',z\mapsto e]$ for
$\alpha,\alpha':T$ and $x:T, z:\alpha=x \vdash P:\Set$ and
$pr:P[x\mapsto \alpha,z\mapsto\refl]$ and $e :\alpha=\alpha'$. We
write $\tr_P\,e\,u: P\,\alpha'$ for transport of $u : P\,\alpha$ along
$e :\alpha=\alpha'$. Sometimes we omit the parameters in subscript. We
also have coercion $\coe\,e\,\alpha:T'$ whenever $e:T=T'$ and the
inverse $e^{-1}:\alpha'=\alpha$ for $e:\alpha=\alpha'$.

In this paper we use the notation of extensional type theory, that is,
after proving an equality $p : \alpha=\alpha'$, in later proofs
depending on this equality we treat it as definitional, thus
$\alpha\equiv\alpha'$ and $p \equiv \refl$. This makes the notation
much lighter, as transports disappear: $\tr\,p\,u\equiv u$ for any
proof $p$. Although we use extensional notation, this can
be translated to intensional type theory extended with functional
extensionality and UIP, as we know from
\cite{hofmann95conservativity, Oury2005,theo}. In our Agda formalisation, we use these axioms
along with rewrite rules \cite{cockx}. A
rewrite rule allows to turn a previously proven (or postulated)
equality into a definitional one. % Whenever Agda encounters a stuck
% term, it looks at the current list of postulated rewrite rules and
% checks whether the term matches the left hand side of one of them. If
% this is the case, it rewrites the term along that rule.
This provides
an alternative way to add inductive types to Agda: one can postulate
the constructors, the eliminator and the computation rules and make
the computation rules definitional by marking them as rewrite rules.

We denote the functional extensionality axiom by $\funext : (\alpha :
A)\ra f\,\alpha = g\,\alpha \ra f = g$. We also use the term $\UIP$
which has type $e = e'$ whenever $e,e':\alpha=\beta$. We write
equational reasoning proofs by writing the proofs of equalities above
the $=$ symbol, e.g. $u \overset{e}{=} u'$ when $e:t = t'$ and
$u\equiv f\,t$, $u'\equiv f\,t'$.

@@

As an example of using the substitution calculus, we prove the functor
laws for terms using equational reasoning.
\begin{alignat*}{15}
  & [\id] && : && t[\id] &&                      && && \hspace{6em} [\circ] && : && t[\sigma\circ\delta] \\
  & && && && && {\rhd\beta_2}^{-1}                && && && && && {\rhd\beta_2}^{-1} \\
  & && = \,\, && \pi_2\,(\id, t[\id]) &&         && && && = \,\, && \pi_2\,(\sigma\circ\delta, t[\sigma\circ\delta]) \\
  & && && && && {\idl}^{-1}                       && && && && && {\idl}^{-1} \\
  & && = \,\, && \pi_2\,(\id\circ\id, t[\id])    && && && && = \,\, && \pi_2\,(\id\circ(\sigma\circ\delta), t[\sigma\circ\delta]) \\
  & && && && && {,\circ}^{-1}                    && && && && && {,\circ}^{-1} \\
  & && = \,\, && \pi_2\,((\id, t)\circ\id) &&    && && && = \,\, && \pi_2\,((\id, t)\circ(\sigma\circ\delta)) \\
  & && && && && \idr                             && && && && && \ass^{-1} \\
  & && = \,\, && \pi_2\,(\id, t) &&              && && && = \,\, && \pi_2\,(((\id, t)\circ\sigma)\circ\delta) \\
  & && && && && {\rhd\beta_2}                    && && && && && {,\circ} \\
  & && = \,\, && t &&                            && && && = \,\, && \pi_2\,((\id\circ\sigma, t[\sigma])\circ\delta) \\
  & && && &&                                     && && && && && && {,\circ} \\
  & && && &&                                     && && && = \,\, && \pi_2\,((\id\circ\sigma)\circ\delta, t[\sigma][\delta]) \\
  & && && &&                                     && && && && && && {\rhd\beta_2} \\
  & && && &&                                     && && && = \,\, && t[\sigma][\delta]
\end{alignat*}

@@

We start by defining $\Con$ as the type of $\mathbb{N}$-algebras,
$\Ty$ as displayed $\mathbb{N}$-algebras, $\Tms$ as
$\mathbb{N}$-homomorphisms and $\Tm$ as sections of displayed
$\mathbb{N}$-algebras, using the same definitions as in section
\ref{sec:naturals}.

@@

\begin{figure}
{\footnotesize
\[
    \begin{tikzpicture}[
node distance = 0pt,
    BC/.style = {
        decorate,
        decoration={calligraphic brace, amplitude=1.2mm,
        pre =moveto, pre  length=0.75pt,
        post=moveto, post length=0.75pt,
        raise=1mm,
        #1},% for mirroring of brace
        very thick,
        pen colour={black},
                  },
%  BC/.default = mirror,
    LN/.style = {inner xsep=4pt, outer sep=0pt},
                        ]
\matrix (m) [matrix of nodes, inner sep=0pt,
             nodes={text depth=0.8ex, text height=1em, %minimum width=5ex,
                    inner ysep=1pt, inner xsep=4pt, outer sep=0pt, anchor=west},
             nodes in empty cells,
             column sep=-\pgflinewidth,
             row sep= -\pgflinewidth,
             ]
{
             & $\blank^\A$   & $\blank^\D$    & $\blank^\M$  & $\blank^\S$  & $\cdot_{\blank}$ & $\rhd_{\blank}$ & ... & $\K_{\blank}$ & ${\K[]}_{\blank}$ & ... & $\Eq_{\blank}$  & ... & \hspace{2em} & $\blank^\C$ & $\blank^\R$ & $\blank^\E$ &  \\
    $\Con$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Ty$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Tms$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Tm$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\cdot$  &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\rhd$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\U$     &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\El$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Pi$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Id$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $\Pim$   &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
    $...$    &              &                &              &              &                 &                &     &         &     &      &  &  &    &   & & &     \\
};
\draw           (m-1-1.south west) -- (m-1-13.south east);
\draw           (m-1-15.south west) -- (m-1-17.south east);
